---
layout: post
title: "OHLC plots with base graphics in R"
date: 2017-08-04
description: ""
category: 
tags: [R, plotting]
---
{% include JB/setup %}


# OHLC Plots With Base Graphics In R
I think OHLC plots are kind of neat, but it seems R caters mostly to scientific data and these are more for financial data. Even so, there are plenty of times that I want to display high-low-mean data, but a boxplot isn't quite right. There are a number of packages that you can download to gain this feature, but oftentimes it is more trouble than it is worth to install a package so you can use one function from it (although [the version from plotly](https://plot.ly/r/ohlc-charts/) looks very cool). Here's a short workaround, using only the base plotting capabilities.  

For data I'll use some wind speed forecasts for a not so-random spot in the Colorado desert. These are based on a 10-day forecast from weatherunderground.com (maybe more on that in the future), so it is 10 rows of predicted min, max, and mean wind speeds.   

```{r}
(winds <- read.csv('winds.csv'))
```

Plotting this table as an OHLC chart is very simple --  I'll just open a blank plot, place my segments on it, then the mean points, and be done:  

```{r}
plot(0,0, type='n', ylim=c(0, max(winds)), xlim=c(0, nrow(winds)-1), xlab='Day into the future', ylab='Wind speed (mph)')
segments(0:(nrow(winds)-1), winds$minwind, 0:(nrow(winds)-1), winds$maxwind)
```

If I want to get fancy, I can write a quick running-average function and draw a line on the plot. Surprisingly, this is not a function in R's base packages either. Of course, you can get this [in many packages](https://stackoverflow.com/questions/743812/calculating-moving-average) if you want to do that instead.  

```{r}
rMu <- function(
    x,        # Vector of values to average
    wl = 3   # Window length: number of frames to average
){
    trail <- wl %/% 2
    lead <- wl %/% 2  - (wl + 1) %% 2 # When wl is even, extra value is behind center
    sapply(1:length(x), function(y) {
        thiswin <- (y-trail):(y+lead)
        vals <- na.omit(x[thiswin[thiswin > 0]])
        sum(vals)/length(vals)
    })
}

plot(0,0, type='n', ylim=c(0, max(winds)), xlim=c(0, nrow(winds)-1), xlab='Day into the future', ylab='Wind speed (mph)')
segments(0:(nrow(winds)-1), winds$minwind, 0:(nrow(winds)-1), winds$maxwind)
points(0:(nrow(winds)-1), winds$avewind, pch='_')
lines(0:(nrow(winds)-1), rMu(winds$avewind, wl=5), lty=3)
```

Of course, if you want to split hairs, this chart only shows three values, and an OHLC chart shows 4. You could add a 4th easily by offsetting your tickmarks.  


```{r}
randoffset <- runif(10, winds$minwind, winds$maxwind)
plot(0,0, type='n', ylim=c(0, max(winds)), xlim=c(0, nrow(winds)-1), xlab='Day into the future', ylab='Wind speed (mph)')
segments(0:(nrow(winds)-1), winds$minwind, 0:(nrow(winds)-1), winds$maxwind)
points(0:(nrow(winds)-1)-0.08, winds$avewind + randoffset, pch='_')
points(0:(nrow(winds)-1)+0.08, winds$avewind, pch='_')
```





















